A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2015-10-28 imaging_demo/ms.tex
Improve explanation for image diff, update figures
From cadd827 (master):
=====================

(*) Add updated image comparison figure, uvwts comparison figure to
the repository and remove older versions

In ms.tex,

(*) Correct minor typos throughtout the paper
(*) Update figures 6, 7, 8, 9 and 12
(*) Recalculate throughput for FX data rates in case of transients
(*) Emphasize the whole point of EPIC
(*) Provide a unified explanation for differences seen with the two
imaging approaches

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts 
consisting of a large number of antennas 
($\Nant\gtrsim 1000$--$\gtrsim 10000$). Since the complexity of traditional 
correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep cost for 
realizing the full imaging potential of these powerful instruments. Through 
our generic and efficient E-field Parallel Imaging Correlator (EPIC), we present 
the first software demonstration of a generalized direct imaging algorithm 
known as the Modular Optimal Frequency Fourier (MOFF) imager. It takes 
advantage of the multiplication-convolution theorem of Fourier transforms. Not 
only does it bring down the cost to $\mathcal{O}(\Nant\log \Nant)$ 
but can also image from irregularly arranged heterogeneous antenna arrays. 
EPIC is highly modular and parallelizable. It is implemented in object oriented 
Python and is publicly available. We have verified the images produced to be 
mathematically identical to those produced using traditional techniques. We have 
also validated our implementation on data observed with the Long Wavelength 
Array. Antenna systems with a dense filling factor consisting of a large number 
of antennas such as LWA, SKA, and HERA will gain a significant advantage by 
deploying EPIC. Inherent availability of calibrated time-domain images on 
timescales roughly equal to the writeout timescale of the digitizer and vastly 
lower bandwidth relative to visibility based systems will make it a prime 
candidate for transient searches of Fast Radio Bursts (FRB) as well as 
planetary and exoplantenary phenomena. 
\end{abstract}

@@

Radio astronomy is entering an era in which interferometers of hundreds to
thousands of individual antennas are needed to achieve desired survey speeds.
Nowhere is this more apparent than at radio frequencies below 1.4 GHz. The study
of the history of hydrogen gas throughout the universe's evolution is pushing
technology development towards arrays of low-cost antennas with large fields of
view and densely packed apertures. Similarly, the search for transient objects
and regular monitoring of the time-dependent sky is driving instruments in the
same direction with the added requirement of fast read-outs. A number of new 
telescopes are under development around the world based on this new paradigm, 
including the Murchison Widefield Array (MWA;\citealt{tin13,bow13}), the 
Precision Array for Probing the Epoch of Reionization (PAPER; \citealt{par10}), 
the Hydrogen Epoch of Reionization Array\footnote{http://reionization.org} 
(HERA), the LOw Frequency ARray (LOFAR;\citealt{van13}), the Canadian Hydrogen 
Intensity Mapping Experiment (CHIME,\citealt{ban14}), the Long Wavelength 
Array (LWA, \citealt{ell13}), and the low frequency component of the Square 
Kilometer Array Low Frequency Aperture Array (SKA1-Low \citealt{mel13}).

This paradigm shift requires a fundamentally new approach to the design of
digital correlators \citep{lon00}. Modern correlators calculate the cross-power
correlation between all antenna pairs in many narrow frequencies, forming
\emph{visibilities}, the traditional fundamental measurement of radio
interferometers. The computational requirements for a modern FX correlator scale
with the number of antenna pairs, or the square of the number of antennas $\sim
\Nant^2$ \citep{bun04}. For this reason traditional correlators have difficulty
scaling to thousands of antennas. As an example, the full HERA correlator for
352 dishes with 200 MHz of bandwidth requires 212 trillion complex multiplies
and adds per second (TMACS). Future arrays with thousands of collecting elements
will require orders of magnitude more computation, making the correlator the
dominant cost.

For certain classes of radio arrays there is an alternative to the FX correlator
that can lower the computational burden by directly performing a spatial fast
Fourier transform (FFT) on the electric fields measured by each antenna in the
array at each time step, removing the cross-correlation step. This relieves the
computational scaling from the harsh $\Nant^2$ to the more gentle envelope of
$\Ngrid\log\Ngrid$, where $\Ngrid$ is the number of grid pixels in the Fourier 
transform \citep[e.g.][]{mor11,teg09,teg10}. This architecture is often referred 
to as a ``direct imaging" correlator because it eliminates the intermediary 
cross-correlation data products of the FX and earlier lag correlators, but 
instead directly forms images from the electric field measurements.

Direct imaging correlators have begun to be explored on deployed arrays including
the Basic Element for SKA Training II (BEST-2) array \citep{fos14}, the Omniscope
\citep{zhe14}, and an earlier incarnation at higher frequencies with the intent
of pulsar timing \citep{oto94, dai00}. However, each of these examples make
assumptions about the redundancy of the array layout, and require the collecting
elements are identical. On the other hand, the MOFF algorithm achieves the same
$\Ngrid \log \Ngrid$ computational scaling without placing any restriction on
antenna placement, can accommodate non-identical beam patterns, and is 
a provably optimal mapping \citep{mor11}. This algorithm uses the antenna beam 
patterns to grid the electric field measurements to a regular grid in the software
holography/A-transpose fashion \citep{mor09,bha08,teg97b} before performing the
spatial FFT. This process has been shown to theoretically produce a data product
identical to images produced from the traditional FX correlator.

Here we present the first software implementation of the MOFF correlator, and
announce the public release of the E-field Parallel Imaging Correlator (EPIC)
code. We begin with a technical description of the algorithm in \S\ref{sec:math}, 
then discuss our particular implementation in \S\ref{sec:software}. We then 
verify the output data quality from our code in \S\ref{sec:verify} by presenting 
simulated images from both the EPIC correlator and comparing to a simulated FX 
correlator. We also demonstrate the performance with real-world data from the 
LWA. In \S \ref{sec:analysis} we analyze the scaling relationships of the 
algorithm. We identify specific array design classes where the EPIC correlator 
is computationally more efficient and in the field of transients, demands 
significantly lesser bandwidth relative to visibility based approaches. We 
conclude and discuss future research prospects in \S\ref{sec:conclusions}.

@@

Interferometers measure {\it visibilities} -- the degree of coherence between
electric fields measured by a pair of antennas \citep{van34,zer38,tho01}.
A visibility, $\widetilde{V}_p$, can be written as:
\begin{align}
  \widetilde{V}_p = &\left\langle \widetilde{E}_a\widetilde{E}_b^\star \right\rangle_t \label{eqn:cc-vis}\hfill\\
  		= &\left\langle \left[ \int {W}_a(\hat{\mathbf{s}})\,E(\hat{\mathbf{s}})\,e^{-i2\pi\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}}\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_a \right] \right. \nonumber\\
		&\qquad \times \left.\left[ \int {W}^\star_b(\hat{\mathbf{s'}})\,E^\star (\hat{\mathbf{s'}})\,e^{i2\pi\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}}\,\dif^2\hat{\mathbf{s'}} + \widetilde{n}^\star_b \right] \right\rangle_t \\
                  =& \iint {W}_a(\hat{\mathbf{s}}) {W}^\star_b(\hat{\mathbf{s'}}) \left\langle E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) \right\rangle_t e^{-i2\pi(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})}\,\dif^2\hat{\mathbf{s}}\,\dif^2\hat{\mathbf{s'}},
\end{align}
where we have brought the time average into the integral under the assumption that 
the aperture illumination pattern does not change over the time-scale of the 
averaging. This expression can be further simplified with the sky brightness, 
$I(\hat{\mathbf{s}})= \left\langle E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) 
\right\rangle_t  \delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the 
antenna pair sky power response function (or the primary beam), 
$B_p(\hat{\mathbf{s}})~\equiv~{W}_a(\hat{\mathbf{s}})~{W}^\star_b(\hat{\mathbf{s}})$. 
The result is the visibility expressed in terms of the sky brightness, the primary 
beam, and uncorrelated noise terms which we group into $\widetilde{n}_p$,
\begin{equation}
\widetilde{V}_p = \int e^{-i2\pi\mathbf{u}_p\!\cdot\,\hat{\mathbf{s}}}\,B_p(\hat{\mathbf{s}})\,I(\hat{\mathbf{s}})\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_p,
\end{equation}
where the baseline coordinate $\mathbf{u}_p=\mathbf{r}_a-\mathbf{r}_b$ is the 
vector separation between the two antennas. This signifies that the visibility 
($\widetilde{V}_p$) measured between a
pair of antennas ($p$) is obtained by the multiplying the sky brightness
$I(\hat{\mathbf{s}})$ by the antenna power response $B_p(\hat{\mathbf{s}})$ and
Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to $uv$
coordinates, which are then sampled at the locations of the antenna spacings (or
baselines), namely, $\mathbf{u}_p$, and added to the receiver noise $n_p$.

This can
be equivalently re-written as:
\begin{align}\label{eqn:software-holography}
  \widetilde{V}_p &= \int \tilde{B}(\mathbf{u}^\prime-\mathbf{u}) \times \left[\int e^{-i2\pi\mathbf{u}.\hat{\mathbf{s}}}\,I(\hat{\mathbf{s}})\,\dif^2\hat{\mathbf{s}}\right]\dif^2\mathbf{u}\vphantom{\int} + n_p,
\end{align}
where, $\tilde{B}(\mathbf{u})$ denotes the $uv$-space antenna power response
obtained by a Fourier transform of $B(\hat{\mathbf{s}})$. Effectively, the
multiplication in image space by $B(\hat{\mathbf{s}})$ has been replaced by a
convolution with $\tilde{B}(\mathbf{u})$ in $uv$-space. This is the software
holographic equivalent of traditional FX correlator output.

@@

Fig.~\ref{fig:MOFF-flowchart} shows the flowchart for MOFF imaging. The
propagated electric fields are shown on the left at different time stamps,
$t_1\ldots t_\textrm{M}$. At each time stamp, the electric fields measured by
antennas are denoted by $E_1(t)\ldots E_\textrm{N}(t)$. The F-engine performs a
temporal Fourier transform on the electric field time-series to obtain electric
field spectra $E_1(f)\ldots E_\textrm{N}(f)$ 
($\widetilde{\mathbf{E}}(\mathbf{a})$ in matrix notation) for each of the 
antennas. Each of the complex antenna gains are calibrated to correct the 
corresponding electric field spectra. These calibrated electric fields are 
gridded using an antenna-based gridding convolution function after which it is 
spatially Fourier transformed and squared to obtain images for every time stamp. 
These images are then time-averaged to obtain the accumulated image 
$I(f)$ ($\mathbf{I}(\hat{\mathbf{s}})$ in matrix notation).
\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_flowchart.eps}
  \caption{A flowchart of MOFF imaging in EPIC. The propagated electric fields
    shown on the left are measured as time-series $E_1(t)\ldots E_\textrm{N}(t)$
    by the antennas which are then Fourier transformed by the F-engine to produce
    electric field spectra $E_1(f)\ldots E_\textrm{N}(f)$. They are calibrated and
    gridded. The gridded electric fields $E_\textrm{g}(f)$ from each time series
    are imaged to produce an images $I_1(f)\ldots I_\textrm{N}(f)$. These images
    are time-averaged to obtain the final image $I(f)$.}
  \label{fig:MOFF-flowchart}
\end{figure}

Fig.~\ref{fig:FX-flowchart} shows the flowchart for a visibility based software 
holographic imaging from a FX correlator. The antenna-based F-engine is identical 
to that in the MOFF processing. The electric field spectra from each antenna are 
then cross-multiplied in the X-engine with those from all other antennas to 
obtain the visibilities $V_\textrm{ij}(f)$ ($\mathbf{m}(\mathbf{v})$ in matrix 
notation). They are calibrated and time-averaged to obtain 
$\langle V_\textrm{ij}(f)\rangle$ which are then gridded and imaged to obtain the 
image $I(f)$. The $I(f)$ obtained from both techniques are identical as explained 
in \S\ref{sec:math}.
\begin{figure}
  \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of visibility based software holographic imaging in EPIC. 
    The FX process flow shares the F-engine with the MOFF process. Following the 
    F-engine, the electric fields pass through the X-engine to obtain 
    visibilities $V_\textrm{ij}(f)$ which are calibrated and time-averaged. Then 
    they are gridded to obtain the gridded visibilities $V_\textrm{g}(f)$ which 
    are then Fourier transformed to obtain the image.
    $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

\par\medskip
\noindent {\bf Temporal Fourier transform}
\par\medskip
\noindent This module is common to the MOFF and visibility based imaging 
techniques. Time samples of electric fields measured by the antenna and 
digitized by the A/D converter is Fourier transformed to generate electric 
field spectra. This step can be parallelized by antennas as shown in 
Fig.~\ref{fig:f-engine}. The output is then fed to either MOFF or visibility 
based imaging pipelines.

@@

In the most general case, this gridding kernel could contain information on the
$w$-projection effect, and even other time-dependent ionospheric effects. For a
stationary antenna array in the absence of any time-dependent effects, this
mapping must only be determined once in the antenna array coordinate frame. The
antenna-to-grid mapping matrix, $\mathbf{M}(\mathbf{r},\mathbf{a})$ is described
as a transformation matrix from the space of measured electric fields by the 
antennas ($\mathbf{a}$) to the antenna array grid denoted by the coordinate 
$\mathbf{r}$. Since each antenna occupies a footprint typically the size of its 
aperture, $\mathbf{M}(\mathbf{r},\mathbf{a})$, which is generally of size
$\Ngrid\times \Nant$, reduces to a sparse block-diagonal matrix
with only $\Nant$ blocks and roughly $N_\textrm{k}$ non-zero entries per
block. This sparse matrix is stored in a Compressed Sparse Row (CSR) format. 
Fig.~\ref{fig:a2g-mapping} illustrates the antenna-to-grid mapping
matrix and the grid containing the mapped aperture footprints of the antennas.

@@

@@

\par\medskip
\noindent {\bf Gridding Convolution}
\par\medskip
\noindent The antenna array aperture illumination over the entire grid,
$\widetilde{\mathbf{W}}(\mathbf{r})$, is obtained by a projection of the 
individual antenna aperture illuminations:
\begin{align}\label{eqn:gridding-convolution}
  \widetilde{\mathbf{W}}(\mathbf{r}) &= \sum_a \widetilde{\mathbf{W}}_a(\mathbf{r}-\mathbf{r}_a) \\
                            &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\mathcal{I}(\mathbf{a}),
\end{align}
where, $\mathcal{I}(\mathbf{a})$ is a row of ones. This is achieved by
efficient multiplication with the sparse matrix created in the antenna-to-grid
mapping process using the sparse matrices module in Python's SciPy package. 
Unless $\widetilde{\mathbf{W}}(\mathbf{r})$ includes time-dependent
effects of the ionosphere or the instrument, it needs to be computed just once
for the entire observation. However, the gridding of electric fields must be
computed at every readout of the electric field spectra,
\begin{align}
  \widetilde{\mathbf{E}}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{E}}(\mathbf{a}).
\end{align}
% The sentence below was already stated earlier.
%In practice the grid is set to be a power of 2 such that the grid spacing is at
%most $\lambda/2$ even at the highest frequency in the observing band.

@@

In MOFF imaging, these are spatially Fourier transformed followed by a squaring
operation at every timestamp for every frequency channel. In visibility based 
imaging, the spatial Fourier transform is performed only once per integration 
timescale and does not include a squaring operation.

@@

\begin{figure*}
  \includegraphics[width=\linewidth]{EPIC-modules.eps}
  \caption{Software architecture of EPIC with core modules, their essential
    attributes and functions. The antenna module forms the fundamental building 
    block. It consists of electric field time-series and spectra and the 
    F-engine that performs a temporal FFT to obtain electric field spectra from 
    the time-series. The interferometer module is made of a pair of antenna 
    modules. Its main function is the X-engine (FX or XF) to produce visibility 
    spectra. The antenna array module is made of all individual antenna modules
    as its components and contains collective properties about the antenna 
    subsystems. Its core function is the creation of antenna-to-grid mapping and 
    that of gridded aperture weights and electric fields. The interferometer 
    array module is very similar in principle to the antenna array module 
    except it operates on cross-correlations and produces gridded visibilities.
    The image module takes gridded electric fields or visibilities and performs
    a two-dimensional spatial FFT (and squares the intermediate image in case 
    of the former) to produce output images. Broadly, the MOFF algorithm is 
    implemented by modules below the horizontal dashed line while the visibility 
    based imaging uses modules above the line. The exact processing pathway 
    implementing the MOFF algorithm is shown in bolded modules.}
  \label{fig:software-modules}
\end{figure*}

@@

The antenna array module consists of all the antenna modules as its attributes
and represents the collective properties of its component antennas. By virtue of 
holding each antenna data independently in their respective modules the F-engine 
for the entire array can be distributed to the F-engines of the component 
antenna modules thus achieving a highly parallelized F-engine while emulating 
real telescope systems.

@@

We elaborate on the data rates required with the MOFF and visibility based
algorithms. This is particularly relevant in the context of radio transient 
detection pipelines.

>>>>

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favoring densely packed array layouts consisting of 
large numbers of antennas ($\Nant\gtrsim 1000$). Since the complexity of 
traditional correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep 
cost for realizing the full imaging potential of these powerful instruments. 
Through our generic and efficient E-field Parallel Imaging Correlator (EPIC), 
we present the first software demonstration of a generalized direct imaging 
algorithm known as the Modular Optimal Frequency Fourier (MOFF) imager. It takes 
advantage of the multiplication-convolution theorem of Fourier transforms. Not 
only does it bring down the cost for dense layouts to 
$\mathcal{O}(\Nant\log_2\Nant)$ but can also image from irregularly arranged 
heterogeneous antenna arrays. EPIC is highly modular and parallelizable, 
implemented in object oriented Python, and publicly available. We have verified 
the images produced to be equivalent to those produced using traditional 
techniques. We have also validated our implementation on data observed with the 
Long Wavelength Array (LWA). Antenna systems with a dense filling factor 
consisting of a large number of antennas such as LWA, the Square Kilometre Array, 
Hydrogen Epoch of Reionization Array, and Canadian Hydrogen Intensity Mapping 
Experiment will gain significant advantage by deploying EPIC. Inherent 
availability of calibrated time-domain images on timescales roughly equal to the 
writeout timescale of the digitizer and vastly lower bandwidth relative to 
visibility based systems will make it a prime candidate for transient searches 
of Fast Radio Bursts (FRB) as well as planetary and exoplanetary phenomena. 
\end{abstract}

@@

Radio astronomy is entering an era in which interferometers of hundreds to
thousands of individual antennas are needed to achieve desired survey speeds.
Nowhere is this more apparent than at radio frequencies below 1.4 GHz. The study
of the history of hydrogen gas throughout the universe's evolution is pushing
technology development towards arrays of low-cost antennas with large fields of
view and densely packed apertures. Similarly, the search for transient objects
and regular monitoring of the time-dependent sky is driving instruments in the
same direction with the added requirement of fast read-outs. A number of new 
telescopes are under development around the world based on this new paradigm, 
including the Murchison Widefield Array (MWA;\citealt{tin13,bow13}), the 
Precision Array for Probing the Epoch of Reionization (PAPER; \citealt{par10}), 
the Hydrogen Epoch of Reionization Array\footnote{http://reionization.org} 
(HERA), the LOw Frequency ARray (LOFAR;\citealt{van13}), the Canadian Hydrogen 
Intensity Mapping Experiment (CHIME,\citealt{ban14}), the Long Wavelength 
Array (LWA, \citealt{ell13}), and the low frequency Square Kilometer Array 
(SKA1-Low \citealt{mel13}).

This paradigm shift requires a fundamentally new approach to the design of
digital correlators \citep{lon00}. Modern correlators calculate the cross-power
correlation between all antenna pairs in many narrow frequencies, forming
\emph{visibilities}, the fundamental measurement of traditional radio
interferometers. The computational requirements for a modern FX correlator scale
with the number of antenna pairs, or the square of the number of antennas $\sim
\Nant^2$ \citep{bun04}. For this reason traditional correlators have difficulty
scaling to thousands of antennas. As an example, the full HERA correlator for
352 dishes with 200 MHz of bandwidth requires 212 trillion complex multiplies
and adds per second (TMACS). Future arrays with thousands of collecting elements
will require orders of magnitude more computation, making the correlator the
dominant cost.

For certain classes of radio arrays there is an alternative to the FX correlator
that can lower the computational burden by directly performing a spatial fast
Fourier transform (FFT) on the electric fields measured by each antenna in the
array at each time step, removing the cross-correlation step. This relieves the
computational scaling from the harsh $\Nant^2$ to the more gentle envelope of
$\Ngrid\log_2\Ngrid$, where $\Ngrid$ is the number of pixels in the Fourier 
transform \citep[e.g.][]{mor11,teg09,teg10}. This architecture is often referred 
to as a ``direct imaging" correlator because it eliminates the intermediary 
cross-correlation data products of the FX and earlier lag correlators, but 
instead directly forms images from the electric field measurements.

Direct imaging correlators have begun to be explored on deployed arrays including
the Basic Element for SKA Training II (BEST-2) array \citep{fos14}, the Omniscope
\citep{zhe14}, and an earlier incarnation at higher frequencies with the intent
of pulsar timing \citep{oto94, dai00}. However, each of these examples make
assumptions about the redundancy of the array layout, and require the collecting
elements are identical. On the other hand, the MOFF algorithm achieves the same
$\Ngrid \log_2 \Ngrid$ computational scaling without placing any restriction on
antenna placement, can accommodate non-identical beam patterns, and is provably 
optimal \citep{mor11}. This algorithm uses the antenna beam patterns to 
grid the electric field measurements to a regular grid in the software 
holography/A-transpose fashion \citep{mor09,bha08,teg97b} before performing the
spatial FFT. This process has been shown to theoretically produce a data product
identical to images produced from traditional visibility based techniques.

Here we present the first software implementation of the MOFF correlator, and
announce the public release of the E-field Parallel Imaging Correlator (EPIC)
code. EPIC is a highly parallel, object oriented Python package that primarily 
implements the MOFF imaging algorithm besides emulating real-life telescopes and 
FX/XF correlators in software, and includes a visibility based imaging technique 
for reference. It is intended to provide a development platform to test different 
imaging approaches, characterize scaling relations and serve as a stepping stone 
for real-life GPU/FPGA based implementation on telescopes.

We begin with a technical description of the algorithm in \S\ref{sec:math}, 
then discuss our particular implementation in \S\ref{sec:software}. We then 
verify the output data quality from our code in \S\ref{sec:verify} by presenting 
simulated images from both the EPIC correlator and comparing to a simulated FX 
correlator. We also demonstrate the performance with real-world data from the 
LWA. In \S \ref{sec:analysis} we analyze the scaling relationships of the 
algorithm. We identify specific array design classes where the EPIC correlator 
is computationally more efficient and in the field of transients, demands 
significantly lesser bandwidth relative to visibility based approaches. We 
conclude and discuss future research prospects in \S\ref{sec:conclusions}.

@@

Interferometers measure {\it visibilities} -- the degree of coherence between
electric fields measured by a pair of antennas \citep{van34,zer38,tho01}.
A visibility, $\widetilde{V}_p$, can be written as:
\begin{align}
  \widetilde{V}_p = &\left\langle \widetilde{E}_a\widetilde{E}_b^\star \right\rangle_t \label{eqn:cc-vis}\hfill\\
  		= &\left\langle \left[ \int {W}_a(\hat{\mathbf{s}})\,E(\hat{\mathbf{s}})\,e^{-i2\pi\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}}\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_a \right] \right. \nonumber\\
		&\qquad \times \left.\left[ \int {W}^\star_b(\hat{\mathbf{s'}})\,E^\star (\hat{\mathbf{s'}})\,e^{i2\pi\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}}\,\dif^2\hat{\mathbf{s'}} + \widetilde{n}^\star_b \right] \right\rangle_t \\
                  =& \iint {W}_a(\hat{\mathbf{s}}) {W}^\star_b(\hat{\mathbf{s'}}) \left\langle E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) \right\rangle_t e^{-i2\pi(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})}\,\dif^2\hat{\mathbf{s}}\,\dif^2\hat{\mathbf{s'}},
\end{align}
where we have brought the time average into the integral under the assumption that 
the aperture illumination pattern does not change over the time-scale of the 
averaging and $\star$ denotes complex conjugation. This expression can be further 
simplified with the sky brightness, 
$I(\hat{\mathbf{s}})= \left\langle E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) 
\right\rangle_t  \delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the 
antenna pair sky power response function (or the directional antenna power pattern), 
$B_p(\hat{\mathbf{s}})~\equiv~{W}_a(\hat{\mathbf{s}})~{W}^\star_b(\hat{\mathbf{s}})$. 
The result is the visibility expressed in terms of the sky brightness, the power 
response, and uncorrelated noise terms which we group into $\widetilde{n}_p$.
\begin{equation}
\widetilde{V}_p = \int e^{-i2\pi\mathbf{u}_p\!\cdot\,\hat{\mathbf{s}}}\,B_p(\hat{\mathbf{s}})\,I(\hat{\mathbf{s}})\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_p,
\end{equation}
where, the baseline coordinate $\mathbf{u}_p=\mathbf{r}_a-\mathbf{r}_b$ is the 
vector separation between the two antennas. This signifies that the visibility 
($\widetilde{V}_p$) measured between a
pair of antennas ($p$) is obtained by the multiplying the sky brightness
$I(\hat{\mathbf{s}})$ by the antenna power response $B_p(\hat{\mathbf{s}})$ and
Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to $uv$
coordinates, which are then sampled at the locations of the antenna spacings (or
baselines), namely, $\mathbf{u}_p$, and added to the receiver noise $n_p$.

This can be equivalently re-written as:
\begin{align}\label{eqn:software-holography}
  \widetilde{V}_p &= \int \widetilde{B}(\mathbf{u}^\prime-\mathbf{u}) \left[\int e^{-i2\pi\mathbf{u}.\hat{\mathbf{s}}}\,I(\hat{\mathbf{s}})\,\dif^2\hat{\mathbf{s}}\right]\dif^2\mathbf{u}\vphantom{\int} + n_p,
\end{align}
where, $\widetilde{B}(\mathbf{u})$ denotes the $uv$-space antenna power response
obtained by a Fourier transform of $B(\hat{\mathbf{s}})$. Effectively, the
multiplication in image space by $B(\hat{\mathbf{s}})$ has been replaced by a
convolution with $\widetilde{B}(\mathbf{u})$ in $uv$-space. This is the software
holographic equivalent of traditional FX correlator output.

@@

Fig.~\ref{fig:MOFF-flowchart} shows the flowchart for MOFF imaging. The
propagated electric fields are shown on the left at different time stamps,
$t_1\ldots t_\textrm{M}$. At each time stamp, the electric fields measured by
antennas are denoted by $E_1(t)\ldots E_\textrm{N}(t)$. The F-engine performs a
temporal Fourier transform on the electric field time-series to obtain electric
field spectra $E_1(f)\ldots E_\textrm{N}(f)$ 
($\widetilde{\mathbf{E}}(\mathbf{a})$ in matrix notation) for each of the 
antennas. Each of the complex antenna gains are calibrated to correct the 
corresponding electric field spectra. These calibrated electric fields are 
gridded using an antenna-based gridding convolution function after which it is 
spatially Fourier transformed and squared to obtain an image cube for every time 
stamp. These images are then time-averaged to obtain the accumulated image 
$I(f)$ ($\mathbf{I}(\hat{\mathbf{s}})$ in matrix notation).
\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_flowchart.eps}
  \caption{A flowchart of MOFF imaging in EPIC. The propagated electric fields
    shown on the left are measured as time-series $E_1(t)\ldots E_\textrm{N}(t)$
    by the antennas $\textrm{A}_1\ldots \textrm{A}_\textrm{N}$ which are then 
    Fourier transformed by the F-engine to produce electric field spectra 
    $E_1(f)\ldots E_\textrm{N}(f)$. They are calibrated and gridded. The gridded 
    electric fields $E_\textrm{g}(f)$ from each time series are imaged to produce 
    images $I_1(f)\ldots I_\textrm{M}(f)$. These images are time-averaged to 
    obtain the final image $I(f)$.}
  \label{fig:MOFF-flowchart}
\end{figure}

Fig.~\ref{fig:FX-flowchart} shows the flowchart for a visibility based software 
holographic imaging from a FX correlator. The antenna-based F-engine is identical 
to that in the MOFF processing. The electric field spectra from each antenna are 
then cross-multiplied in the X-engine with those from all other antennas to 
obtain the visibilities $V_\textrm{ij}(f)$ ($\mathbf{m}(\mathbf{v})$ in matrix 
notation). They are calibrated and time-averaged to obtain 
$\langle V_\textrm{ij}(f)\rangle$ which are then gridded and imaged to obtain the 
image $I(f)$. The $I(f)$ obtained from both techniques are theoretically 
identical as explained in \S\ref{sec:math}.
\begin{figure}
  \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of visibility based software holographic imaging in EPIC. 
    The FX process flow shares the F-engine with the MOFF process. Following the 
    F-engine, the electric fields pass through the X-engine to produce 
    visibilities $V_\textrm{ij}(f)$ which are calibrated and time-averaged. Then 
    they are gridded to obtain the gridded visibilities $V_\textrm{g}(f)$ which 
    are then Fourier transformed to obtain the image, $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

@@

In the most general case, this gridding kernel could contain information on
$w$-projection effects, and other time-dependent ionospheric effects. For a
stationary antenna array in the absence of any time-dependent effects, this
mapping must only be determined once in the antenna array coordinate frame. The
antenna-to-grid mapping matrix, $\mathbf{M}(\mathbf{r},\mathbf{a})$ is described
as a transformation matrix from the space of measured electric fields by the 
antennas ($\mathbf{a}$) to the antenna array grid denoted by the coordinate 
$\mathbf{r}$. Since each antenna occupies a footprint typically the size of its 
aperture, $\mathbf{M}(\mathbf{r},\mathbf{a})$, which is generally of size
$\Ngrid\times \Nant$, reduces to a sparse block-diagonal matrix
with only $\Nant$ blocks and roughly $N_\textrm{k}$ non-zero entries per
block. This sparse matrix is stored in a Compressed Sparse Row (CSR) format. 
Fig.~\ref{fig:a2g-mapping} illustrates the antenna-to-grid mapping
matrix and the grid containing the mapped aperture footprints of the antennas.

@@

\par\medskip
\noindent {\bf Temporal Fourier transform}
\par\medskip
\noindent This module is common to the MOFF and visibility based imaging 
techniques. Time samples of electric fields measured by the antenna and 
digitized by the A/D converter is Fourier transformed to generate electric 
field spectra. This step can be parallelized across antennas as shown in 
Fig.~\ref{fig:f-engine}. The output is then fed to either MOFF or visibility 
based imaging pipelines.

@@

\par\medskip
\noindent {\bf Gridding Convolution}
\par\medskip
\noindent The antenna array aperture illumination over the entire grid,
$\widetilde{\mathbf{W}}(\mathbf{r})$, is obtained by a projection of the 
individual antenna aperture illuminations:
\begin{align}\label{eqn:gridding-convolution}
  \widetilde{\mathbf{W}}(\mathbf{r}) &= \sum_a \widetilde{\mathbf{W}}_a(\mathbf{r}-\mathbf{r}_a) \\
                            &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\mathcal{I}(\mathbf{a}),
\end{align}
where, $\mathcal{I}(\mathbf{a})$ is a row of ones. This is achieved by
efficient multiplication with the sparse matrix created in the antenna-to-grid
mapping process using the sparse matrices module in Python SciPy package. 
Unless $\widetilde{\mathbf{W}}(\mathbf{r})$ includes time-dependent
effects of the ionosphere or the instrument, it needs to be computed just once
for the entire observation. However, the gridding of electric fields must be
computed at every readout of the electric field spectra,
\begin{align}
  \widetilde{\mathbf{E}}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{E}}(\mathbf{a}).
\end{align}
% The sentence below was already stated earlier.
%In practice the grid is set to be a power of 2 such that the grid spacing is at
%most $\lambda/2$ even at the highest frequency in the observing band.

@@

In MOFF imaging, these are spatially Fourier transformed followed by a squaring
operation at every time stamp for every frequency channel. In visibility based 
imaging, the spatial Fourier transform is performed only once per integration 
timescale and does not include a squaring operation.

@@

\begin{figure*}
  \includegraphics[width=\linewidth]{EPIC-modules.eps}
  \caption{Software architecture of EPIC with core modules, their essential
    attributes and functions. The antenna module forms the fundamental building 
    block. It consists of electric field time-series and spectra and the 
    F-engine that performs a temporal FFT to obtain electric field spectra from 
    the time-series. The interferometer module is made of a pair of antenna 
    modules. Its main function is the X-engine (FX or XF) to produce visibility 
    spectra. The antenna array module is made of all individual antenna modules
    as its components and contains collective properties about the antenna 
    subsystems. Its core function is the creation of antenna-to-grid mapping, 
    gridded aperture weights and electric fields. The interferometer array 
    module is very similar in principle to the antenna array module except it 
    operates on cross-correlations and produces gridded visibilities. The image 
    module takes gridded electric fields or visibilities and performs a 
    two-dimensional spatial FFT (and squares the intermediate image in case 
    of the former) to produce output images. Broadly, the MOFF algorithm is 
    implemented by modules below the horizontal dashed line while the visibility 
    based imaging uses modules above the line. The exact processing pathway 
    implementing the MOFF algorithm is shown in bolded modules.}
  \label{fig:software-modules}
\end{figure*}

@@

The antenna array module consists of all the antenna modules as its attributes
and represents the collective properties of its component antennas. By virtue of 
holding each antenna data independently in their respective modules, the F-engine 
for the entire array can be distributed to the F-engines of the component 
antenna modules thus achieving a highly parallelized F-engine while emulating 
real telescope systems.

@@

We elaborate on the data rates required with the MOFF and visibility based
algorithms. This is particularly relevant in the context of radio transient 
detection.
